---
title: "R Coding Exercise (Module 3)"
output: 
  html_document:
    theme: flatly
    toc: FALSE
---
<br>

## Introduction

The following exercise is for [Module 3 in Dr. Andreas Handel's MADA Course](https://andreashandel.github.io/MADAcourse/Assessment_Coding.html).

The main portion of this exercises utilizes the Gapminder dataset to examine the relationship between infant mortality and life expectancy as well as population size and life expectancy in African countries. It conducts some data processing, creates a few plots, and adds a simple linear models of the data. The appendix, added by fellow group member Joe Martin, adds regression lines to the figures and creates dataframes for the simple linear models.

---

## Required Packages
The following R packages are required for this exercise:

 * dslabs: for loading the Gapminder dataset
 * dplyr: for data wrangling
 * ggplot2: for data plotting
 * broom: for linear model outputs
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

#to use the gapminder data
library("dslabs") 

#to select certain columns of data
library("dplyr") 

#to plot data
library(ggplot2)

#to create dataframes of linear model outputs
library(broom)

#to center all plot titles
theme_update(plot.title = element_text(hjust = 0.5))
```


---

## Checking data
```{r}
#look at help file for gapminder data
utils::help(gapminder)

#overview of data structure
utils::str(gapminder)

#summary of data
summary(gapminder)

#determine object type of gapminder data
class(gapminder)
```

---

## Processing data
First we will subset the Gapminder data to include only African countries. Then we will make two new subsets: one with only infant mortality and life expectancy data and another with only population and life expectancy data.
```{r}
# Goal: create new object that selects only African countries
africadata <- gapminder[gapminder$continent == "Africa", ]

#overview of Africa data structure
utils::str(africadata)

#summary of Africa data
summary(africadata)

### new subset of Africa data that only includes infant mortality and life expectancy
africa_im_le <- africadata %>%
  select(infant_mortality, life_expectancy)

#overview of infant mortality and life expectancy data
utils::str(africa_im_le)

#summary of infant mortality and life expectancy data
summary(africa_im_le)

### new subset of Africa data that only includes population and life expectancy
africa_pop_le <- africadata %>%
  dplyr::select(population, life_expectancy)

#overview of population and life expectancy data
utils::str(africa_pop_le)

#summary of population and life expectancy data
summary(africa_pop_le)
```

---

## Plotting
Create plots for each of the new subsets. We're not too concerned with aesthetics just yet - we just want to visualize the relationships.
```{r}
#plot life expectancy as a function of infant mortality 
ggplot2::ggplot(data = africa_im_le, aes(x = infant_mortality, y = life_expectancy)) +
           geom_point() +
           scale_size_manual(values = 2) +
           ggtitle("Life expectancy as a function of infant mortality") +
           labs(x = "Infant Mortality", y = "Life Expectancy")

#plot life expectancy as a function of population size
ggplot2::ggplot(data = africa_pop_le, aes(x = population, y = life_expectancy)) +
          geom_point() +
          scale_size_manual(values = 2) +
          scale_x_continuous(trans = 'log2') +
          ggtitle("Life expectancy as a function of population") +
          labs(x = "log(Population)", y = "Life Expectancy")
```

---

## More data processing
The Gapminder data includes data for different years for individual countries. First, we will identify years with missing data, and then create a subset for a year with complete data (2000).
```{r}
### Identify missing infant mortality data
#make a list of missing infant mortality data
africadata_missing <- africadata[is.na(africadata$infant_mortality), ]

#identify years of missing infant mortality data
unique(africadata_missing$year)

### Create subset of Africa data only in year 2000
africadata_2000 <- africadata[africadata$year == "2000", ]

#overview of Africa data in 2000
utils::str(africadata_2000)

#summary of Africa data in 2000
summary(africadata_2000)
```

---

## More plotting
Replicate the plots above, but for the 2000 subset data. Again, not worried about aesthetics here - just focused on visualizing the relationship.
```{r}
#new subset of Africa data in 2000 that only includes infant mortality and life expectancy
africa_im_le_2000 <- africadata_2000 %>%
  dplyr::select(infant_mortality, life_expectancy)

#overview of infant mortality and life expectancy data in 2000
utils::str(africa_im_le_2000)

#summary of infant mortality and life expectancy data in 2000
summary(africa_im_le_2000)

#plot of life expectancy as a function of infant mortality in 2000
ggplot2::ggplot(data = africa_im_le_2000, aes(x = infant_mortality, y = life_expectancy)) +
          geom_point() +
          scale_size_manual(values = 2) +
          ggtitle("Life expectancy as a function of infant mortality") +
          labs(x = "Infant Mortality", y = "Life Expectancy")

### new subset of Africa data in 2000 that only includes population and life expectancy
africa_pop_le_2000 <- africadata_2000 %>%
  dplyr::select(population, life_expectancy)

#overview of population and life expectancy data in 2000
utils::str(africa_pop_le_2000)

#summary of population and life expectancy data in 2000
summary(africa_pop_le_2000)

#plot of life expectancy as a function of population in 2000
ggplot2::ggplot(data = africa_pop_le_2000, aes(x = population, y = life_expectancy)) +
          geom_point() +
          scale_size_manual(values = 2) +
          scale_x_continuous(trans = 'log2') +
          ggtitle("Life expectancy as a function of population") +
          labs(x = "log(Population)", y = "Life Expectancy")
```

---

## A simple fit
We are using the lm function to fit life expectancy as the outcome, and infant mortality and population size as predictors (respectively) using the year 2000 subsetted data.
```{r}
#linear model with life expectancy as outcome, infant mortality as predictor (using 2000 data)
fit1 <- stats::lm(life_expectancy ~ infant_mortality, data = africa_im_le_2000)

#summary of fit 1
summary(fit1)

#linear model with life expectancy as outcome, population size as predictor (using 2000 data)
fit2 <- stats::lm(life_expectancy ~ population, data = africa_pop_le_2000)

#summary of fit 2
summary(fit2)
```

---

## Conclusion
In examining the two linear models, we do not have enough evidence to determine that population size is a significant predictor of life expectancy in African countries in the year 2000. However, there is sufficient evidence to suggest a negative correlation between infant mortality rate and life expectancy in African countries in the year 2000. For each additional infant death per 1,000, the life expectancy decreases by 0.18916 years (SE = 5.459e-08, t = 0.505, p < 0.001).

---

## Appendix A - Group Work (Added by Joe Martin)


### Regression Lines
```{r}
#for infant mortality and life expectancy
im_le2k <- africa_im_le_2000 %>%
         ggplot2::ggplot(aes(x=infant_mortality, y=life_expectancy))+
                   geom_point()+
                   geom_smooth(method="lm")

im_le2k

#for population and life expectancy
pop_le_lm2k <- africa_pop_le_2000 %>% 
               ggplot2::ggplot(aes(x=log10(population), y=life_expectancy))+
                         geom_point()+
                         geom_smooth(method="lm")

pop_le_lm2k
```

---

### Linear Model Summaries
```{r}
#for life expectancy and infant mortality
broom::tidy(fit1)

#for population and life expectancy
broom::tidy(fit2)
```
