---
title: "R Coding Exercise (Module 3)"
output: 
  html_document:
    theme: journal
    toc: FALSE
---
<br>

The first portion of this exercises utilizes the Gapminder dataset to examine the relationship between infant mortality and life expectancy as well as population size and life expectancy in African countries.

<br>

## Libraries and other formatting
```{r}
#to use the gapminder data
library("dslabs") 

#to select certain columns of data
library("dplyr") 

#to plot data
library(ggplot2)

#to center all plot titles
theme_update(plot.title = element_text(hjust = 0.5))
```

<br>

## Checking data
```{r}
#look at help file for gapminder data
help(gapminder)

#overview of data structure
str(gapminder)

#summary of data
summary(gapminder)

#determine object type of gapminder data
class(gapminder)
```

<br>

## Processing data
```{r}
### Goal: create new object that selects only African countries
africadata <- gapminder[gapminder$continent == "Africa", ]

#overview of Africa data structure
str(africadata)

#summary of Africa data
summary(africadata)

### new subset of Africa data that only includes infant mortality and life expectancy
africa_im_le <- africadata %>%
  select(infant_mortality, life_expectancy)

#overview of infant mortality and life expectancy data
str(africa_im_le)

#summary of infant mortality and life expectancy data
summary(africa_im_le)

### new subset of Africa data that only includes population and life expectancy
africa_pop_le <- africadata %>%
  select(population, life_expectancy)

#overview of population and life expectancy data
str(africa_pop_le)

#summary of population and life expectancy data
summary(africa_pop_le)
```

<br>

## Plotting
```{r}
#plot life expectancy as a function of infant mortality 
ggplot(data = africa_im_le, aes(x = infant_mortality, y = life_expectancy)) +
  geom_point() +
  scale_size_manual(values = 2) +
  ggtitle("Life expectancy as a function of infant mortality") +
  labs(x = "Infant Mortality", y = "Life Expectancy")

#plot life expectancy as a function of population size
ggplot(data = africa_pop_le, aes(x = population, y = life_expectancy)) +
  geom_point() +
  scale_size_manual(values = 2) +
  scale_x_continuous(trans = 'log2') +
  ggtitle("Life expectancy as a function of population") +
  labs(x = "log(Population)", y = "Life Expectancy")
```

<br>

## More data processing
```{r}
### Identify missing infant mortality data
#make a list of missing infant mortality data
africadata_missing <- africadata[is.na(africadata$infant_mortality), ]

#identify years of missing infant mortality data
unique(africadata_missing$year)

### Create subset of Africa data only in year 2000
africadata_2000 <- africadata[africadata$year == "2000", ]

#overview of Africa data in 2000
str(africadata_2000)

#summary of Africa data in 2000
summary(africadata_2000)
```

<br>

## More plotting
```{r}
#new subset of Africa data in 2000 that only includes infant mortality and life expectancy
africa_im_le_2000 <- africadata_2000 %>%
  select(infant_mortality, life_expectancy)

#overview of infant mortality and life expectancy data in 2000
str(africa_im_le_2000)

#summary of infant mortality and life expectancy data in 2000
summary(africa_im_le_2000)

#plot of life expectancy as a function of infant mortality in 2000
ggplot(data = africa_im_le_2000, aes(x = infant_mortality, y = life_expectancy)) +
  geom_point() +
  scale_size_manual(values = 2) +
  ggtitle("Life expectancy as a function of infant mortality") +
  labs(x = "Infant Mortality", y = "Life Expectancy")

### new subset of Africa data in 2000 that only includes population and life expectancy
africa_pop_le_2000 <- africadata_2000 %>%
  select(population, life_expectancy)

#overview of population and life expectancy data in 2000
str(africa_pop_le_2000)

#summary of population and life expectancy data in 2000
summary(africa_pop_le_2000)

#plot of life expectancy as a function of population in 2000
ggplot(data = africa_pop_le_2000, aes(x = population, y = life_expectancy)) +
  geom_point() +
  scale_size_manual(values = 2) +
  scale_x_continuous(trans = 'log2') +
  ggtitle("Life expectancy as a function of population") +
  labs(x = "log(Population)", y = "Life Expectancy")
```

<br>

## A simple fit
```{r}
#linear model with life expectancy as outcome, infant mortality as predictor (using 2000 data)
fit1 <- lm(life_expectancy ~ infant_mortality, data = africa_im_le_2000)

#summary of fit 1
summary(fit1)

#linear model with life expectancy as outcome, population size as predictor (using 2000 data)
fit2 <- lm(life_expectancy ~ population, data = africa_pop_le_2000)

#summary of fit 2
summary(fit2)
```

<br>

## Conclusion
In examining the two linear models, we do not have enough evidence to determine that population size is a significant predictor of life expectancy in African countries in the year 2000. However, there is sufficient evidence to suggest a negative correlation between infant mortality rate and life expectancy in African countries in the year 2000. For each additional infant death per 1,000, the life expectancy decreases by 0.18916 years (SE = 5.459e-08, t = 0.505, p < 0.001).